---
title: "MetaWeave architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MetaWeave architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Vision

MetaWeave reconstructs spatial ecological structure from incomplete information. Its central abstraction is not a raster or a stochastic block model, but a local ecological system: co-occurring species and inferred relationships among them.

## Object pipeline

1. A `distribution_collection` answers where species can occur.
2. An `assemblage` represents local coexistence without assuming interactions.
3. An `inference_model` maps an assemblage to an `ecological_network`.
4. Simulations turn probabilistic networks into possible observations.
5. Metrics summarize networks without knowing how they were inferred.
6. A `spatial_result` maps those summaries back to geography.

## Pluggable inference

The first generic backend is `block_model()`, which assigns interaction probabilities from latent block pairs. Its row and column group names are supplied by the user. A trait backend can use exactly the same contract:

```{r}
library(metaweave)

trait_predict <- function(model, assemblage) {
  producers <- assemblage$species$producers
  consumers <- assemblage$species$consumers
  p <- outer(producers, consumers, Vectorize(function(i, j) model$parameters$score(i, j)))
  dimnames(p) <- list(producers, consumers)
  new_ecological_network(p, "probability", row_group = "producers", column_group = "consumers")
}

trait_model <- new_inference_model(
  trait_predict,
  name = "trait matching",
  parameters = list(score = function(palm, mammal) 0.5)
)
```

Downstream simulation, summaries, and maps do not change when the model changes.

## Design boundaries

MetaWeave coordinates ecological inference. It does not replace GIS, species-distribution modelling, specialist network statistics, or visualization packages. This boundary keeps its model interface stable while allowing those tools to evolve independently.
