---
title: "Getting started with metaweave"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with metaweave}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`metaweave` combines species distributions with an interaction model to infer
local ecological networks. This tutorial uses invented species and probabilities;
no downloads or external data files are needed. In a real study, the probabilities
could come from an independently fitted model or expert knowledge. They are
inputs here, not quantities estimated from the presence data.

## Define interaction probabilities

Rows represent plants and columns represent animals. An entry of 0.8 means an
80% modelled probability of interaction when both species are present. It is
neither an observed interaction count nor a species occurrence probability.

```{r}
library(metaweave)
probabilities <- matrix(
  c(0.8, 0.2, 0.3, 0.7), nrow = 2, byrow = TRUE,
  dimnames = list(c("plant_a", "plant_b"), c("animal_x", "animal_y"))
)
model <- probability_matrix_model(probabilities, "plants", "animals")
```

## Infer one local network

An assemblage lists species present at one location. Group names must match
those supplied to the model. This model selects relevant rows and columns from
the supplied matrix; it does not refit the probabilities locally.

```{r}
site <- new_assemblage(list(plants = c("plant_a", "plant_b"),
                            animals = "animal_x"))
network <- infer_network(site, model)
network$matrix
summarize_network(network)
```

There are two possible plant-animal pairs. Their probabilities sum to 1.1,
the expected number of links. An expectation can be fractional: it describes an
average over possible outcomes, rather than a count of observed interactions.

## Apply the model across a small landscape

Each raster layer represents one species and each cell represents one location.
Nonzero values indicate presence; zero or missing values indicate absence.
All groups must share the same grid and coordinate reference system. Species
layer names must match the interaction matrix.

```{r}
grid <- create_standard_grid(c(0, 2, 0, 2), resolution = 1)
plants <- terra::rast(list(grid, grid))
animals <- terra::rast(list(grid, grid))
names(plants) <- rownames(probabilities)
names(animals) <- colnames(probabilities)
terra::values(plants) <- cbind(c(1, 1, 0, 1), c(1, 0, 1, 1))
terra::values(animals) <- cbind(c(1, 1, 1, 0), c(0, 1, 1, 1))
result <- run_spatial_inference(
  distributions = list(plants = plants, animals = animals),
  model = model, min_species = 1
)
result$result$index
```

The workflow identifies local species, infers their networks, summarizes each
network and maps the summaries to the original grid. `min_species = 1` retains
cells with at least one species in each group. The index contains cell IDs,
coordinates and summaries. Individual networks are in `result$result$networks`.

```{r map, fig.width=5, fig.height=4}
terra::plot(result$spatial$data[["expected_links"]], main = "Expected links per cell")
```

This map shows expected link counts under the supplied probabilities and presence
data. It does not show observed interactions or uncertainty intervals.

## Other models

- `block_model()` uses supplied probabilities between species blocks.
- `maxent_model()` generates directed network ensembles subject to link-count or
  degree constraints. Its help page explains assumptions and sampling limits.
- `new_inference_model()` connects a custom prediction function to the workflow.
- `simulate_interaction_counts()` allocates a fixed total of counts in proportion
  to matrix entries. These counts are not independent binary links.

See function help pages for inputs and returned objects. Use
`citation("metaweave")` for the software citation.

