---
title: "Maximum-entropy network inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maximum-entropy network inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

MetaWeave's first maximum-entropy backend generates binary food webs under
explicit structural constraints and averages them into marginal link
probabilities. It implements the minimally biased constrained-ensemble idea of
Banville, Gravel and Poisot (2023), but uses exact fixed-link sampling or a
degree-preserving swap chain instead of their SVD-entropy simulated annealing.

## Directed food web

Rows are consumers and columns are resources. The same species pool can occupy
both roles.

```{r}
library(metaweave)

web <- new_assemblage(
  list(species = c("algae", "grazer", "fish", "bird")),
  cell_id = 1
)

model <- maxent_model(
  row_group = "species",
  constraint = "connectance",
  connectance = 0.25,
  ensemble_size = 500,
  self_links = FALSE,
  seed = 42
)

probability_web <- infer_network(web, model)
round(probability_web$matrix, 2)
```

Every ensemble member has exactly `round(0.25 * 12) = 3` links because four
species provide 12 directed non-self dyads. Each matrix entry is the fraction of
ensemble members containing that interaction.

## Fixed degree sequences

Named degree vectors can preserve the number of resources used by each consumer
and the number of consumers using each resource.

```{r}
degrees <- c(algae = 1, grazer = 1, fish = 1, bird = 1)

degree_model <- maxent_model(
  row_group = "species",
  constraint = "degree_sequence",
  out_degree = degrees,
  in_degree = degrees,
  ensemble_size = 250,
  self_links = FALSE,
  seed = 42,
  burn_in = 500,
  thin = 50
)

degree_web <- infer_network(web, degree_model)
round(degree_web$matrix, 2)
```

The degree sampler first finds a feasible adjacency matrix and then applies
checkerboard swaps that preserve every row and column sum. It approximates a
uniform distribution over feasible matrices. Convergence diagnostics and the
SVD-entropy objective are intentionally deferred to a later backend revision.

## Reproducibility

A non-null seed creates a local random stream and restores R's previous random
state after inference. For spatial collections, the cell identifier is added to
the seed, giving cells distinct but repeatable ensembles.

## Reference

Banville F, Gravel D, Poisot T (2023). "What constrains food webs? A maximum
entropy framework for predicting their structure with minimal biases." *PLOS
Computational Biology*, 19, e1011458.
