---
title: "MetaWeave: rationale, methods, and simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MetaWeave: rationale, methods, and simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Rationale

Species distributions tell us which interactions are possible in a place, but
not which interactions are likely. A regional interaction model tells us which
species may interact, but not where they co-occur. MetaWeave combines these two
sources of information to reconstruct local ecological networks.

The package separates three questions:

1. **Where are species present?** Distribution layers define local assemblages.
2. **How are co-occurring species connected?** An inference model assigns
   interaction probabilities.
3. **What might an observed sample look like?** Simulation converts a
   probability network into interaction counts.

This separation allows the ecological model to change without rewriting the
spatial workflow.

## Core workflow

MetaWeave uses a small set of objects and transformations:

| Object | Meaning |
|---|---|
| `distribution_collection` | Named spatial distributions for one group |
| `assemblage` | Species co-occurring in one spatial unit |
| `inference_model` | Rule for assigning interactions |
| `ecological_network` | Local interaction matrix and its meaning |
| `network_collection` | Local networks linked to spatial cells |
| `spatial_result` | Network summaries mapped back to geography |

The main pipeline is:

```text
distributions -> assemblages -> inference model -> local networks
              -> summaries -> spatial rasters
```

## Inference methods

MetaWeave currently provides two general inference backends.

### Block model

`block_model()` assumes that species belong to ecological groups or latent
blocks. The probability of interaction between species (i) and (j) is

\[
P(Y_{ij}=1)=\Theta_{z_i,z_j},
\]

where (z_i) and (z_j) are their block memberships and \(\Theta\) is the
block-to-block probability matrix. Memberships and \(\Theta\) may come from an
external stochastic block model, expert-defined guilds, or another grouping
method.

### Species probability matrix

`probability_matrix_model()` accepts a named species-by-species probability
matrix directly. This is useful for trait models, machine-learning predictions,
Bayesian posterior means, or ensemble predictions that do not correspond to one
block partition.

Both methods return the same network class, so simulation, summaries, and maps
are shared.

## Simulated example

The following example creates a small landscape with six producer species and
five consumer species. The distributions are artificial, so the full analysis
is reproducible without external data.

```{r libraries}
library(metaweave)
library(terra)

set.seed(42)
```

### Create a landscape and distributions

```{r distributions}
grid <- create_standard_grid(
  extent = c(0, 6, 0, 4),
  resolution = 1,
  crs = "EPSG:4326"
)

make_range <- function(template, center_x, center_y, spread) {
  xy <- terra::crds(template)
  distance <- sqrt(
    (xy[, 1] - center_x)^2 +
      (xy[, 2] - center_y)^2
  )

  layer <- terra::rast(template)
  terra::values(layer) <- ifelse(distance <= spread, 1, NA)
  layer
}

producer_parameters <- data.frame(
  species = paste0("producer_", 1:6),
  x = c(1.0, 2.0, 3.0, 4.0, 5.0, 3.0),
  y = c(1.0, 3.0, 2.0, 3.0, 1.0, 3.5),
  spread = c(2.0, 2.2, 1.8, 2.0, 1.8, 2.4)
)

consumer_parameters <- data.frame(
  species = paste0("consumer_", 1:5),
  x = c(1.5, 2.5, 4.0, 5.0, 3.0),
  y = c(2.5, 1.0, 2.0, 3.0, 3.5),
  spread = c(2.0, 1.8, 2.2, 1.8, 2.4)
)

producers <- terra::rast(lapply(seq_len(nrow(producer_parameters)), function(i) {
  make_range(
    grid,
    producer_parameters$x[i],
    producer_parameters$y[i],
    producer_parameters$spread[i]
  )
}))
names(producers) <- producer_parameters$species

consumers <- terra::rast(lapply(seq_len(nrow(consumer_parameters)), function(i) {
  make_range(
    grid,
    consumer_parameters$x[i],
    consumer_parameters$y[i],
    consumer_parameters$spread[i]
  )
}))
names(consumers) <- consumer_parameters$species
```

Each raster layer represents one species. A value of `1` means presence and
`NA` means absence.

### Define a block-level interaction model

Producers and consumers are assigned to two blocks. The matrix `theta` states
that matching blocks interact more strongly than mismatched blocks.

```{r block-model}
producer_lookup <- data.frame(
  species = names(producers),
  guild = rep(c("producer_A", "producer_B"), each = 3)
)

consumer_lookup <- data.frame(
  species = names(consumers),
  guild = c("consumer_A", "consumer_A", "consumer_B",
            "consumer_B", "consumer_B")
)

theta <- matrix(
  c(
    0.80, 0.20,
    0.25, 0.75
  ),
  nrow = 2,
  byrow = TRUE,
  dimnames = list(
    c("producer_A", "producer_B"),
    c("consumer_A", "consumer_B")
  )
)

model <- block_model(
  row_lookup = producer_lookup,
  column_lookup = consumer_lookup,
  theta = theta,
  row_group = "producers",
  column_group = "consumers"
)
```

### Reconstruct local networks

`run_spatial_inference()` finds local assemblages, infers their networks,
calculates summaries, and maps those summaries to the original grid.

```{r inference}
result <- run_spatial_inference(
  distributions = list(
    producers = producers,
    consumers = consumers
  ),
  model = model,
  min_species = 1
)

head(result$result$index)
```

The standard summaries are:

- `row_richness` and `column_richness`: local species richness;
- `possible_links`: number of co-occurring species pairs;
- `expected_links`: sum of their interaction probabilities; and
- `mean_probability`: average probability across possible pairs.

```{r maps, fig.show='hold'}
plot(
  result$spatial$data[[
    c("row_richness", "column_richness", "expected_links")
  ]],
  main = c(
    "Producer richness",
    "Consumer richness",
    "Expected links"
  )
)
```

Spatial variation emerges from both species turnover and the interaction model.
Two cells with equal richness can have different expected links if their species
belong to different blocks.

## Simulating interaction observations

A probability network describes ecological expectation. Field data are finite
samples from that expectation. `simulate_interaction_counts()` distributes a
fixed number of interaction events among species pairs in proportion to their
inferred probabilities.

We select the local network with the largest number of possible links.

```{r focal-network}
focal_id <- which.max(result$result$index$possible_links)
focal_network <- result$result$networks[[focal_id]]

round(focal_network$matrix, 2)
```

Simulate a small and a large survey:

```{r sampling}
counts_50 <- simulate_interaction_counts(
  focal_network,
  size = 50,
  seed = 1
)

counts_500 <- simulate_interaction_counts(
  focal_network,
  size = 500,
  seed = 1
)

sum(counts_50)
sum(counts_500)
```

Larger samples usually represent the underlying relative interaction weights
more closely. We can demonstrate this across repeated simulations.

```{r simulation-analysis}
target <- focal_network$matrix / sum(focal_network$matrix)

sampling_error <- function(sample_size, replicates = 100) {
  errors <- vapply(seq_len(replicates), function(i) {
    observed <- simulate_interaction_counts(
      focal_network,
      size = sample_size,
      seed = 1000 + i
    )
    observed <- observed / sum(observed)
    mean(abs(observed - target))
  }, numeric(1))

  data.frame(
    sample_size = sample_size,
    mean_absolute_error = mean(errors),
    error_sd = stats::sd(errors)
  )
}

simulation_results <- do.call(
  rbind,
  lapply(c(25, 50, 100, 250, 500), sampling_error)
)

simulation_results

plot(
  simulation_results$sample_size,
  simulation_results$mean_absolute_error,
  type = "b",
  log = "x",
  xlab = "Number of sampled interactions",
  ylab = "Mean absolute error"
)
```

This analysis measures sampling error, not whether the inference model is
biologically correct. Increasing sample size reduces observation noise but does
not correct a misspecified interaction model.

## Standardizing a network metric

Networks sampled with different effort are difficult to compare. One simple
approach is to simulate each network at the same sample size and then calculate
a metric. Here we use observed connectance.

```{r standardized-metric}
observed_connectance <- function(count_matrix) {
  sum(count_matrix > 0) / length(count_matrix)
}

connectance_100 <- rarefied_network_metric(
  prob_mat = focal_network$matrix,
  target_size = 100,
  metric = observed_connectance,
  n_per_level = 100,
  seed = 99
)

connectance_100
```

This value is the average connectance expected after 100 sampled interaction
events. The same procedure can be applied to every local network or replaced by
a specialist rarefaction method.

## Using a species-level probability matrix

Some models return probabilities for individual species pairs rather than a
single block partition. MetaWeave accepts these predictions directly.

```{r probability-model}
species_probabilities <- theta[
  producer_lookup$guild,
  consumer_lookup$guild,
  drop = FALSE
]

dimnames(species_probabilities) <- list(
  producer_lookup$species,
  consumer_lookup$species
)

matrix_model <- probability_matrix_model(
  probability_matrix = species_probabilities,
  row_group = "producers",
  column_group = "consumers",
  name = "species-level predictions"
)

matrix_result <- run_spatial_inference(
  distributions = list(
    producers = producers,
    consumers = consumers
  ),
  model = matrix_model,
  min_species = 1
)
```

Because this example expands the same block model, its results match the block
backend. In practice, the matrix could contain ensemble, trait-based, or
machine-learning predictions.

## Interpretation and limits

MetaWeave reconstructs plausible local networks conditional on its inputs. A
predicted link is not a confirmed observation. Results inherit uncertainty and
bias from:

- species distributions and their spatial resolution;
- incomplete regional interaction data;
- the chosen inference model and fitted parameters; and
- variation in sampling effort.

The current simulation functions represent finite sampling from a fixed local
probability matrix. They do not yet propagate uncertainty in species ranges,
block membership, model parameters, or environmental responses. Those sources
should be handled through repeated model inputs or future uncertainty-aware
backends.

The main value of MetaWeave is therefore comparative and modular: the same
spatial assemblages can be evaluated under different ecological hypotheses,
and their consequences can be compared across space using a shared workflow.
