---
title: "Reconstructing palm-mammal networks with MetaWeave"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reconstructing palm-mammal networks with MetaWeave}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette reconstructs local palm-mammal networks in four steps: clean the
species distributions, match species across datasets, fit a regional network
model, and project that model into local communities.

## 1. Load packages and data

Change `project_dir` if the project is moved. Keeping one root path makes the
analysis easier to reproduce on another computer.

```{r data}
library(metaweave)
library(terra)
library(dplyr)
library(cassandRa)

project_dir <- paste0(
  "C:/Users/gabri/Documents/PhD/",
  "Network_reconstruction_under_data_scarsity/MetaWeave/MetaWeave"
)

palm_grid_occ <- readRDS(file.path(
  project_dir,
  "DATA/clean_datasets/palm_ranges_grid_stack.RDS"
))

mammal_grid_occ <- readRDS(file.path(
  project_dir,
  "DATA/clean_datasets/mammal_ranges_grid_stack.RDS"
))

palm_traits <- read.csv(file.path(
  project_dir,
  "Data/01_species_traits/final_palm_trait.csv"
))

mammal_traits <- read.csv(file.path(
  project_dir,
  "Data/01_species_traits/final_mammal_trait.csv"
))

palm_mammal_frugivory <- read.csv(file.path(
  project_dir,
  paste0(
    "DATA/02_species_interactions/PalmDryadRepo/",
    "Munozetal2019/PalmFrugDatasetOCT2018.csv"
  )
)) |>
  filter(
    frugClass == "MAMMAL",
    biogeographicRegion == "Neotropics"
  )
```

The trait tables are loaded for future trait-based models. The example below
uses species identities and observed interactions.

## 2. Clean species ranges

Species names must use the same format in every dataset. Duplicate raster
layers are merged because they describe separate polygons for the same species.

```{r clean-ranges}
standardize_species <- function(x) {
  x <- trimws(tolower(as.character(x)))
  x <- gsub("[^[:alnum:]]+", "_", x)
  gsub("^_+|_+$", "", x)
}

merge_duplicate_ranges <- function(x) {
  groups <- factor(names(x), levels = unique(names(x)))

  merged <- terra::tapp(
    x,
    index = groups,
    fun = "sum",
    na.rm = TRUE
  )

  merged <- terra::ifel(merged > 0, 1, NA)
  names(merged) <- levels(groups)
  merged
}

remove_empty_ranges <- function(x) {
  limits <- terra::global(x, c("min", "max"), na.rm = TRUE)
  keep <- !is.na(limits$min) & !is.na(limits$max)
  x[[which(keep)]]
}

names(palm_grid_occ) <- standardize_species(names(palm_grid_occ))
names(mammal_grid_occ) <- standardize_species(names(mammal_grid_occ))

palm_grid_occ <- palm_grid_occ |>
  merge_duplicate_ranges() |>
  remove_empty_ranges()

mammal_grid_occ <- mammal_grid_occ |>
  merge_duplicate_ranges() |>
  remove_empty_ranges()

stopifnot(
  !anyDuplicated(names(palm_grid_occ)),
  !anyDuplicated(names(mammal_grid_occ))
)
```

## 3. Match ranges and interactions

The interaction table uses `PALM` for producers and `FRUGIVORE` for consumers.
Only species present in both the regional interaction data and range rasters are
retained.

```{r match-species}
interaction_data <- palm_mammal_frugivory |>
  transmute(
    producer = standardize_species(PALM),
    consumer = standardize_species(FRUGIVORE)
  ) |>
  filter(
    !is.na(producer),
    !is.na(consumer),
    producer != "",
    consumer != ""
  )

model_producers <- intersect(
  names(palm_grid_occ),
  unique(interaction_data$producer)
)

model_consumers <- intersect(
  names(mammal_grid_occ),
  unique(interaction_data$consumer)
)

stopifnot(
  length(model_producers) > 0,
  length(model_consumers) > 0
)

producer_ranges <- palm_grid_occ[[model_producers]]
consumer_ranges <- mammal_grid_occ[[model_consumers]]

stopifnot(compareGeom(
  producer_ranges,
  consumer_ranges,
  stopOnError = FALSE
))

cat("Matched palms:", length(model_producers), "\n")
cat("Matched mammals:", length(model_consumers), "\n")
```

## 4. Build an observed-metaweb baseline

The baseline treats every recorded regional interaction as present. It is useful
for testing the spatial workflow before fitting a probabilistic model.

```{r observed-model}
observed_interactions <- interaction_data |>
  filter(
    producer %in% model_producers,
    consumer %in% model_consumers
  ) |>
  distinct(producer, consumer) |>
  mutate(observed = 1)

theta_observed <- interaction_table_to_matrix(
  data = observed_interactions,
  row_col = "producer",
  column_col = "consumer",
  value_col = "observed",
  fill = 0
)

producer_lookup <- data.frame(
  species = rownames(theta_observed),
  guild = rownames(theta_observed)
)

consumer_lookup <- data.frame(
  species = colnames(theta_observed),
  guild = colnames(theta_observed)
)

observed_model <- block_model(
  row_lookup = producer_lookup,
  column_lookup = consumer_lookup,
  theta = theta_observed,
  row_group = "producers",
  column_group = "consumers"
)
```

Project the regional metaweb into every cell where at least one palm and one
mammal occur.

```{r observed-spatial}
observed_result <- run_spatial_inference(
  distributions = list(
    producers = producer_ranges,
    consumers = consumer_ranges
  ),
  model = observed_model,
  min_species = 1
)

head(observed_result$result$index)

plot(
  observed_result$spatial$data[[
    c(
      "row_richness",
      "column_richness",
      "expected_links",
      "mean_probability"
    )
  ]],
  main = c(
    "Palm richness",
    "Mammal richness",
    "Recorded links",
    "Local connectance"
  )
)
```

## 5. Fit a stochastic block model

`cassandRa` groups species by their interaction patterns. MetaWeave then uses
the fitted group memberships and block probabilities to infer local networks.

```{r fit-sbm}
regional_web <- xtabs(
  ~ producer + consumer,
  data = interaction_data |>
    filter(
      producer %in% model_producers,
      consumer %in% model_consumers
    )
) |>
  as.matrix()

regional_web <- regional_web[
  rowSums(regional_web) > 0,
  colSums(regional_web) > 0,
  drop = FALSE
]

set.seed(123)

cass_fit <- regional_web |>
  cassandRa::CreateListObject() |>
  cassandRa::FitSBM(n_SBM = 10, G = NULL)

best_sbm <- cass_fit$SBM1
theta_sbm <- best_sbm$Omega_rs

producer_block_names <- paste0(
  "producer_block_",
  seq_len(nrow(theta_sbm))
)

consumer_block_names <- paste0(
  "consumer_block_",
  seq_len(ncol(theta_sbm))
)

rownames(theta_sbm) <- producer_block_names
colnames(theta_sbm) <- consumer_block_names

producer_lookup <- data.frame(
  species = rownames(regional_web),
  guild = producer_block_names[best_sbm$SB_H]
)

consumer_lookup <- data.frame(
  species = colnames(regional_web),
  guild = consumer_block_names[best_sbm$SB_W]
)

sbm_model <- block_model(
  row_lookup = producer_lookup,
  column_lookup = consumer_lookup,
  theta = theta_sbm,
  row_group = "producers",
  column_group = "consumers"
)
```

## 6. Reconstruct and map local SBM networks

The output contains one probability network per eligible cell, a cell index,
and rasters of network summaries.

```{r spatial-sbm}
sbm_result <- run_spatial_inference(
  distributions = list(
    producers = producer_ranges[[producer_lookup$species]],
    consumers = consumer_ranges[[consumer_lookup$species]]
  ),
  model = sbm_model,
  min_species = 1
)

head(sbm_result$result$index)

plot(
  sbm_result$spatial$data[[
    c(
      "row_richness",
      "column_richness",
      "possible_links",
      "expected_links",
      "mean_probability"
    )
  ]],
  main = c(
    "Palm richness",
    "Mammal richness",
    "Possible links",
    "Expected SBM links",
    "Mean SBM probability"
  )
)
```

`expected_links` is the sum of species-pair probabilities in each cell. The
observed baseline describes known links, while the SBM also assigns probability
to interactions that were not recorded regionally.

## 7. Optional ensemble predictions

If `cassandRa::FitSBM()` returns a normalized species-level probability matrix,
MetaWeave can use it directly. This preserves ensemble predictions without
forcing them into one block partition.

```{r ensemble}
ensemble_probabilities <- cass_fit$SBM_ProbsMat
dimnames(ensemble_probabilities) <- dimnames(regional_web)

stopifnot(
  all(is.finite(ensemble_probabilities)),
  all(ensemble_probabilities >= 0),
  all(ensemble_probabilities <= 1)
)

ensemble_model <- probability_matrix_model(
  probability_matrix = ensemble_probabilities,
  row_group = "producers",
  column_group = "consumers",
  name = "cassandRa SBM ensemble"
)

ensemble_result <- run_spatial_inference(
  distributions = list(
    producers = producer_ranges[[rownames(ensemble_probabilities)]],
    consumers = consumer_ranges[[colnames(ensemble_probabilities)]]
  ),
  model = ensemble_model,
  min_species = 1
)
```

